---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

# define formula
mdl_formula <- bf(Distance ~ Sem * Phon)

# get priors
get_prior(mdl_formula, sim)

# specify priors
my_priors <- c(prior(normal(100, 40), class  = 'Intercept'),
               prior(normal(0, 40), class  = 'b'))

# check predictor levels
contrasts(sim$Sem) 
contrasts(sim$Phon)
  # intercept: SemDifferent * PhonDifferent
```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

# run bayesian model
sim_mdl <- brm(mdl_formula,
               family = gaussian(),
               prior = my_priors,
               data = sim,
               chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1,
                  seed = 16)

```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

# check model fit
pp_check(sim_mdl, ndraws = 100)
  #observed data has gentle right tail, square-root transformation to fix
sim <- sim |>
  mutate(sqrt_distance = sqrt(Distance))

#re-specify model with transformed Distance variable
my_priors_sqrt <- c(prior(normal(10, 4), class  = 'Intercept'),
                    prior(normal(0, 4), class  = 'b'))

sim_mdl_sqrt <- brm(sqrt_distance ~ Sem * Phon,
               family = gaussian(),
               prior = my_priors_sqrt,
               data = sim,
               chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1,
                  seed = 16)

# check model fit again
pp_check(sim_mdl_sqrt, ndraws = 100)
  # fit is better
```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

# check model specifications
sim_mdl_sqrt
  #rhat and ESS are good

# check mixing of chains
mcmc_plot(sim_mdl_sqrt, type = 'trace')
  # chains look good

```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

# examine model output
sim_mdl_sqrt
  # credible intervals for slope coefficients (including interaction term) overlap with zero, error is fairly large.

# visualize posteriors

sim_mdl_sqrt |> 
  spread_draws(b_SemSimilar) |> 
  ggplot(aes(x = b_SemSimilar)) +
   stat_halfeye(fill = 'darkorchid3') +
   geom_vline(xintercept = 0, lty = 'dashed') +
   labs(x = "Effect of Semantic Similarity",
        y = "") +
   theme_minimal()

sim_mdl_sqrt |> 
  spread_draws(b_PhonSimilar) |> 
  ggplot(aes(x = b_PhonSimilar)) +
   stat_halfeye(fill = 'darkorchid3') +
   geom_vline(xintercept = 0, lty = 'dashed') +
   labs(x = "Effect of Phonetic Similarity",
        y = "") +
   theme_minimal()

sim_mdl_sqrt |> 
  spread_draws(`b_SemSimilar:PhonSimilar`)|> 
  ggplot(aes(x = `b_SemSimilar:PhonSimilar`)) +
   stat_halfeye(fill = 'darkorchid3') +
   geom_vline(xintercept = 0, lty = 'dashed') +
   labs(x = "Interaction between Semantic and Phonetic Similarity",
        y = "") +
   theme_minimal()

# calculate probability of direction for predictors
sim_mdl_sqrt |>
  spread_draws(b_SemSimilar, b_PhonSimilar, `b_SemSimilar:PhonSimilar`) |>
  select(b_SemSimilar, b_PhonSimilar, `b_SemSimilar:PhonSimilar`) |>
  p_direction()

# calculate Bayes factor
# extract posteriors
posteriors <- sim_mdl_sqrt |>
  spread_draws(b_SemSimilar, b_PhonSimilar, `b_SemSimilar:PhonSimilar`) |>
  select(b_SemSimilar, b_PhonSimilar, `b_SemSimilar:PhonSimilar`)

#extract priors
sim_mdl_priors <- brm(sqrt_distance ~ Sem * Phon,
               family = gaussian(),
               prior = my_priors_sqrt,
               sample_prior = 'only',
               data = sim,
               chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1,
                  seed = 16)

priors <- sim_mdl_priors |>
  spread_draws(b_SemSimilar, b_PhonSimilar, `b_SemSimilar:PhonSimilar`) |>
  select(b_SemSimilar, b_PhonSimilar, `b_SemSimilar:PhonSimilar`)

bayesfactor_parameters(posteriors, priors, direction = "two-sided", null = 0)
  # inconclusive evidence, anecdotal to moderate in favour of the null  
```

## (g) Write up your entire analysis:

Posterior distributions of the model indicate that square-root-transformed Distance decreases by an average of -0.58 \[95% Credible Interval: -1.29 - 0.11\] when changing from the Semantically Different to the Semantically Similar condition. Conversely, when changing from the Phonologically Different to the Phonologically Similar condition, square-root-transformed Distance increases by an average of 0.54 \[95% Credible Interval: -0.10 - 1.22\]. The interaction between the Semantically Similar and Phonologically Similar condition additionally decreases the square-root-transformed Distance by an average of -0.20 \[95% Credible Interval: -1.14 - 0.78\].

For the effect of Semantic Similarity, 94.75% of the posterior samples are negative, indicating that positive values are plausible but unlikely (given the data, the model, and priors). Similarly, for the effect of Phonological Similarity, 94.96% of the posterior samples are positive, indicating that negative values are plausible but unlikely. For the interaction effect between Semantic and Phonological Similarity, 66.35% of the posterior samples are negative, indicating that we cannot make confident claims about the direction of the effect.

Subsequently, we calculate the Savage-Dickey density ratio [@DickeyLientz1970-The-Weighted-Li; @WagenmakersLodewyckx2010-Bayesian-hypoth; @roettgermeaningful], i.e., we relate the evidence (the probability of the posterior distribution at point-zero) for the model based on the priors only (i.e., before seeing the data) to the evidence for the model based on both priors and likelihood (i.e., after seeing the data) [e.g., @WetzelsGrasman2010-An-encompassing]. The Savage-Dickey method lets us assess evidence for and against a null hypothesis using Bayes factor. Here, the Bayes factor indicates the degree by which the mass of the posterior distribution has shifted further away from the point null value (relative to the prior distribution). Our analysis yields Bayes factors of 0.35 and 0.30 for the effect of Semantic Similarity and Phonological Similarity respectively, indicating that the probability of the Similarity effect being point null has increased somewhat, suggesting anecdotal to moderate evidence in favour of the null hypothesis that there is no Similarity effect on the Distance measure. For the interaction effect between the two types of similarity, the analysis yields a Bayes factor of 0.14, indicating that the probability of the interaction effect being point null has increased, suggesting moderate evidence in favour of the null hypothesis that there is no interaction effect. [@lee-wagenmakers_Bayesian_book_2014].

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
